---
title: "PSET B"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this problem set you will investigate variation of student achievement in mathematics within and between a sample of 160 schools. These are a subset of data from the National Center for Education Statistics ([NCES](https://nces.ed.gov)) "High School and Beyond" study ([HSB](https://nces.ed.gov/surveys/hsb/)) that followed a national cohort of high school seniors in the US in 1980. This dataset is used extensively in the Raudenbush and Bryk (2002) text and in limited cases of our textbook. While the dataset is old, we will use these data for two reasons. First, the multilevel structure and questions the dataset was designed to address remain relevant for more recent studies and datasets (e.g., see [here for some recent NCES data collections that may be useful](https://nces.ed.gov/surveys/)). Second, because these data are used in textbooks, it provides the opportunity to replicate the textbook analyses if you choose to do so, which can be a helpful way to connect the `R` code and output with the conceptual presentation of the models.

In the problem set, some steps are primarily for instructional purposes and others are intended to replicate more substantive prior analyses of the data.

**Data file:** hsb.RDS (obtained from the `merTools` package in R: https://rdrr.io/cran/merTools/man/hsb.html) 

**Variables:**

- **schid**: School ID variable
- **minority**: a dummy variable representing student race/ethnicity (0=White, 1=non-White)
- **female**: a dummy variable representing student gender (1=female; 0=male)
- **ses**: A composite measure of social class provided by HSB. In the current sample, mean is approximately 0 and SD is approximately 0.8.
- **mathach**: A mathematics test given in the senior year (mean = 12.7, sd = 6.9; for more info see here)
- **size**: total student enrollment in the school
- **sector**: a dummy variable (1=Catholic school; 0=public school)

## Load Packages

```{r, message=FALSE, warning=FALSE}
library(haven)     # read in data files
library(sjPlot)    # regression tables
library(vtable)    # summary tables
library(lme4)      # estimate MLM
library(lmerTest)  # hypothesis testing for MLM
library(tidyverse) # data management
```

## Load Data

```{r}
hsb <- readRDS(file = "hsb.RDS")
```

## Variable definitions and inclusivity

The dataset uses the term "minority" to identify students' racial or ethnic identities, consistent with the original study. However, it is important for us to use current terminology that is more inclusive to describe constructs such as racial and ethnic identities. Rather than identify students in the dataset as "minority" or not, we will identify students as either "students of color" or not (here, White students). This is in accordance with current guidance on inclusive language from the American Psychological Association (APA; see [here](https://www.apa.org/about/apa/equity-diversity-inclusion/language-guidelines#)).

To reflect this change, rename the variable `minority`:

```{r}
hsb <- rename(hsb, soc=minority)
```

I have used `soc` as an abbreviation for "student of color" and will use this language where relevant. The variable values remain the same - thus, `soc=0` indicates a White student while `soc=1` indicates a student of color. Ideally we would have a more comprehensive set of racial/ethnic identities in the dataset, but these are not available in the data.

Note that student gender identity is also represented only as a binary (female or not). Although gender identity is not fully represented by this binary, again, there is no more detailed information available in the data. 

Hopefully, these limitations serve as a reminder to allow for more comprehensive identities to be represented in data that you collect.

## Question 1

*Create a table of summary statistics for the level 1 variables: soc, female, ses, and mathach. The table should include at least the following statistics for each variable: number of observations, mean, standard deviation, minimum, and maximum.*

```{r}

```

## Question 2

*Create a table of summary statistics for the following level 2 variables: size, sector, average math achievement, percent students of color, average ses. The table should include at least the following statistics for each variable: number of observations, mean, standard deviation, minimum, and maximum. (Note: creating this table requires you to create a school-level dataset so that you can summarize these values at the school level.)*

```{r}

```

## Question 3

*Estimate an "empty" random-intercept model that allows you to estimate the ICC of math achievement across schools (use `REML=FALSE` in `lmer()`). The statistical model, where $Y$ is math achievement, is:*

Level 1:

$$Y_{ij} =\beta_{0j}+e_{ij}$$
Level 2:
$$\beta_{0j}=\gamma_{00}+u_{0j}$$
Random effects:
$$u_{0j} \sim N(0,\sigma^2_{u0}), \space e_{ij}\sim N(0,\sigma^2)$$

*Create a table of results using `tab_model()`.*

```{r}

```

## Question 4

*Write 1 sentence interpreting each of the following statistics:* $\gamma_{00}$, $\sigma_e^2$, $\sigma^2_{u0}$ *(note in `tab_model()` $\tau_{00}=\sigma^2_{u0}$).*

## Question 5

*What is the estimated ICC of student math scores? Briefly interpret this estimated value. Based on your results so far, is there evidence that MLM might be needed to analyze these data? How do you know?*

## Question 6

*Calculate the variance of observed* **average** *math test scores across schools. (HINT: you calculated the standard deviation of observed average math test scores in question 2 above.) How does this estimate of the variance in school mean test scores compare to the estimate of the variance in school means from your model in Question #3? Why do you think these values differ?*

```{r}

```

## Question 7

*Estimate two different regression models:*

a. *an OLS regression model predicting math achievement from students’ family SES.*
b. *a random-intercept model predicting math achievement from students’ family SES that includes school random intercepts (use `REML=FALSE`).*

*Create a nicely formatted regression table showing both model results side by side.*

```{r}

```

## Question 8

*For each model in Question 7, write 1 sentence interpreting each of the following estimates: model intercept and coefficient for SES. Also write 1 sentence interpreting $\sigma^2_{u0}$ from the random intercept model.*

## Question 9

*Briefly comment on what you notice about similarities and differences between the two models estimated in Question 7.*

## Question 10

*Create a new version of the student-level SES variable called "ses_gmc" by centering the `ses` variable at the grand mean in the dataset. (NOTE: the mean of the `ses` variable in the data is already very close to 0, but is not EXACTLY 0. After you do this step, the mean of "ses_gmc" should be exactly 0.) Calculate the mean of your new "ses_gmc" variable to be sure the mean is 0.*

## Question 11

*Estimate a random intercept model predicting math achievement from student ses (level 1 variable, grand mean centered) and school sector (level 2 variable). Use `REML=FALSE`. The model in statistical notation is, where $y$ is math achievement, $X$ is SES, and $W$ is school sector:*

Level 1:

$$y_{ij}=\beta_{0j}+\beta_{1j}(X_{ij}-\bar X)+e_{ij}$$

Level 2:

$$\beta_{0j}=\gamma_{00}+\gamma_{10}(W_j)+u_{0j}$$
$$\beta_{1j}=\gamma_{10}$$

Random effects:

$$u_{0j} \sim N(0,\sigma^2_{u0}), \space e_{ij}\sim N(0,\sigma^2)$$

*Use `tab_model()` to produce a nicely formatted table of the results.*

```{r}

```

## Question 12

*Write 1 sentence interpreting each of the following estimated parameters: $\gamma_{00}$, $\gamma_{01}$, $\gamma_{10}$, $\sigma^2_{u0}$.*

## Question 13

*Briefly explain 1 or 2 substantive conclusions you can draw from the model results in Question 11.*

## Question 14

*What is an example of a multilevel data structure in your field of research? This can by a hypothetical example or a real example you have come across in your own work or reading. Be sure to explain what the outcome ($Y$) variable is as well as what the level 1 units and level 2 (or higher) clusters are.*
