---
title: "Estimating Random Intercept Model in R Part 1"
subtitle: "EDUC 7456"
author: "Benjamin Shear"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This document reviews the R code needed to estimate random intercept models in `R`. The document uses the `popular.sav` example dataset from the Hox et al. (2018) textbook.

## Load Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(vtable)
library(lme4)
library(sjPlot)
library(arm)
library(jtools)
library(insight)
```

## Load Data

In the example here we will use the `popular2.sav` SPSS format dataset from Hox et al. (2018). See Appendix E for discussion of the dataset.

```{r}
dat_popular <- read_spss(file = "popular2.sav")
head(dat_popular) 
```

### Convert factor to numeric

The variable `sex` is a labelled variable. This will sometimes prevent the variable from being summarized as a binary indicator variable. Here I create a new indicator (dummy) variable called `female` equal to 1 if a student's sex is "girl" in the original data and equal to 0 if the student's sex is "boy".

```{r}
dat_popular$female <- ifelse(dat_popular$sex==1,1,0)
```

## Summarize Data

With a multilevel dataset we need to summarize the data at each level of interest to understand the structure.

### Full Dataset

We will use the `sumtable` function from the `vtable` package again to create summary tables. For now we will only ask the function to summarize the following variables:

- pupil
- class
- extrav
- female
- texp
- popular

```{r}
sumtable(data = dat_popular, vars = c("pupil", "class", "extrav", "female", "texp", "popular"))
```

### Summarize Level-1 Variables

For a summary of level-1 variables we can make the same table, but remove the `class` and `texp` variables, which are technically "level-2" variables. They do not vary across students within the same class.

```{r}
sumtable(data = dat_popular, vars = c("pupil", "extrav", "female", "popular"))
```

### Calculate Level-2 Variables

At level-2 we want to summarize the `class` variable and the `texp` variable. To facilitate this, we will create a new dataset in R that has only 100 rows - 1 for each class. For the `texp` variable I will calculate the class mean; however, the `texp` variable doesn't vary within classes so the mean is the same thing as "the" value for that class. I will also ask R to calculate the number of students in each class. Although this is not a variable we will use in the models later, it is very useful information to summarize.

To do the summary, I use the "pipe" operator `%>%` and the `group_by()` and the `summarise()` functions. We will use these often.

```{r}
level2_data <- dat_popular %>%
  group_by(class) %>%  # group by class ID
  summarise(
    MNtexp = mean(texp), # the MNtexp will be the mean of texp in each class
    Nstu = n(),           # Nstu is the number of rows in each class
    mnpop=mean(popular)
  ) %>% ungroup()
```

### Summarize Level-2 Variables

```{r}
sumtable(level2_data)
```

## Empty Random Intercept (RI) Model

The "empty" RI model allows us to estimate the between-class and within-class variance components of the `popular` rating variable. Statistically the model is:

$$popular_{ij}=\gamma_{00}+u_{0j}+e_{ij}$$

$$u_{0j} \sim N(0,\sigma^2_{u_0})$$

$$e_{ij} \sim N(0,\sigma^2_e)$$

### Make Picture of Empty RI Model

We can make graph to show this model using `ggplot()`:

```{r, fig.height=4, fig.width=12}
dat_popular %>%
  ggplot(aes(x=class, y=popular)) +
  geom_hline(yintercept = mean(dat_popular$popular), color="black", lty=2) +
  geom_jitter(alpha=0.75, color="grey", width=0.05) +
  stat_summary(
    fun = "mean",
    geom = "point",
    col = "red",
    shape = 15,
    size = 4,
    alpha = 0.65
  ) +
  scale_x_continuous(breaks = seq(1,100,5)) +
  scale_y_continuous(limits = c(0,10), breaks = seq(0,10,2)) +
  theme_bw()
```

### Estimate with lmer()

We estimate the model using `lmer()`.

```{r}
ri_model <- lmer(popular ~ 1 + (1 | class), data = dat_popular)
```

### Print Results

The simplest way to "see" the results is to use the `summary()` function, although it is a lot of output.

```{r}
summary(ri_model)
```

### How to "get" Variance Components

For now, we are most interested in the variance components. We can have R give those to us in an easier to use way using `VarrCorr()`.

```{r}
VarCorr(ri_model)
```

Or a slightly more detailed version by converting to a data.frame.

```{r}
data.frame(VarCorr(ri_model))
```

But we can also use the [insight](https://easystats.github.io/insight/index.html) package functions.

```{r}
get_variance_intercept(ri_model)
get_variance_residual(ri_model)
```

### Calculate ICC

We can calculate the ICC as:

$$\hat {ICC}=\frac{\hat\sigma^2_{u_0}}{\hat\sigma^2_{u_0}+\hat \sigma^2_e}$$

You could manually calculate this. Using the R objects we can calculate:

Option 1 using `VarCorr()`

```{r}
varcomps <- data.frame(VarCorr(ri_model))
var_btwn <- varcomps[1,4]
var_wthn <- varcomps[2,4]
(var_btwn)/(var_btwn+var_wthn)
```

Option 2 using `insight` functions

```{r}
var_btwn <- get_variance_intercept(ri_model)
var_wthn <- get_variance_residual(ri_model)
(var_btwn)/(var_btwn+var_wthn)
```

Option 3: most of the functions below!

## More Ways to Print Results

Here are three useful ways to summarize the results of a `lmer` object. They are useful for different contexts:

- [tab_model()](https://strengejacke.github.io/sjPlot/articles/tab_model_estimates.html) in the `sJPlot` package is great for making nice tables that you could include in a paper. Very customizable, and can be exported to MS Word and Excel.
- [display()](https://rdrr.io/cran/arm/#vignettes) in the `arm` package is useful for a compact, quick summary to look at results that is less messy than the `summary()` function. This is mostly useful for a quick look at a model, for example while doing analysis or in class.
- [summ()](https://jtools.jacob-long.com/articles/summ.html) in the `jtools` package is somewhere in between. It is a bit more detailed and neater than `display()` but it isn't really "publication-ready" the way `tab_model()` results (almost) are.

### tab_model() in `sjPlot`

Default:

```{r}
tab_model(ri_model)
```

Some customization:

```{r}
tab_model(ri_model,
          dv.labels = "Model Title",
          show.ci=FALSE,
          show.se=TRUE,
          string.se="SE",
          string.est="Est.")
```

### display() in `arm`

```{r}
arm::display(ri_model, digits=3)
```

### summ() in `jtools`

```{r}
summ(ri_model)
```

