---
title: "week 2 regression Practice"
author: "Kate Little"
date: "`r Sys.Date()`"
format: html_document
---

```{r setup}
setwd("C:/Users/littkate/Desktop/EDUC7456")

library(vtable)
library(sjPlot)
library(readxl)
library(tidyverse)

dee24 <- read_excel(path = "regression example/Chronic_Absenteeism_202310.xlsx", range = "A1:O52")

# subset to variables of interest -----

wk02_data <- select(dee24, ST, State, CAR_1819, CAR_2122, COVID19_Rate, REMOTE_2020)

# drop states with missing data -----

wk02_data <- drop_na(wk02_data)

```


# RQ1 
Did states with higher prevalence of remote instruction in 2021 
experience larger increases in chronic absenteeism rates 
from 2019 to 2022?

```{r q1, echo = TRUE}

# y is CAR_2122- CAR_1819
# x1 = REMOTE_2020

# create new var representing change in CAR (chronic absenteesim rate)
wk02_data<-wk02_data%>%
            mutate(CAR_CHANGE = CAR_2122-CAR_1819)

# sumtable is part of vtable package
sumtable(wk02_data)
```

There are 41 states with nonmissing data

Lets make a model and plot on scatter plot
```{r plot, echo= TRUE}
# use lm to create linear model
# read ~ as 'regress on'
change_remote<-lm(CAR_CHANGE ~ REMOTE_2020, data = wk02_data)

plot(wk02_data$REMOTE_2020, wk02_data$CAR_CHANGE)
abline(change_remote, col = "blue")

# show the stats of the model using tab model

# tab_model is part of sjplot
# only works in html
tab_model(change_remote)

```

# Interpretation:

## Intercept: 
The predicted change in CAR from 18/19 to 21/22 for a state whose students spent 0% of the time remotely would be 10.2 units.

## Slope: 
a. For a 0.1 change in time spent learning remotely, one could predict an 1.153% change in absenteeism. 
b. For every 10% of remote learning time in 20/21 the expected difference in absenteeism from 18/19-21/22 would be 1.153 precentage points.

## Rsquared: 
21.5% of the variation in increase in CAR is explained by the prevalence of remote learning in 20/21.



